---
title: "Steps for Visualing Data"
author: "Vanuatu National Statistics Office"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: '2'
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load required libraries, include=FALSE}
library(knitr) # Nice tables
library(kableExtra) # Extra nice tables
```

```{r functions, include=FALSE}
prettyTable <- function(table){
  
  # Use the kable function to create a nicer formatted table
  kable(table) %>%
  
    # Set the format
    kable_styling(bootstrap_options="striped", # Set the colour of rows
                  full_width=FALSE, # Make the table not stretch to fit the page
                  position="left") %>% # Position the table on the left
  
    # Make the table scrollable
    scroll_box(height = "400px")
}
```

## Introduction

The world today is filled with data and it becomes imperative that we analyse it properly to gain meaningful insights. Data Visualisation is a vital tool that can unearth possible crucial insights from data. If the results of an analysis are not visualised properly, it will not be communicated effectively to the desired audience. In this tutorial, we will learn how to analyze and display data using R statistical language. R is extremely flexible and easy to use when it comes to creating visualisations. 

We'll be using the `carInfo-Prices_27-05-20.csv` dataframe, this is the dataset that we cleaned and processed in the previous tutorial. The current tutorial was created using Rmarkdown - a powerful tool to help us create formatted using R code. If you're interested in using Rmarkdown, there is a great tutorial [here](https://rmarkdown.rstudio.com/lesson-1.html).

## Setting the working directory

Next, we'll set the working directory - usually the folder on our computer where we have stored our data. We can use the following code for this:

```{r set working directory}
# Note the name of our working directory
workingDirectory <- file.path("/Users/hugopigott/Dropbox/2. Work/Vanuatu National Statistics Office/Reproducible Analytical Pipelines/RTraining")

# Set the working directory
setwd(workingDirectory)
```


## Loading required libraries

The power of R is that millions of people are using it. R comes with a large set of `base` functions that we'll be using. Sometimes though it is useful to use some additional libraries.  We can install and load libraries with the following code to visualise our data:

```{r installing libraries}
install.packages("dplyr") # We'll use this package to manipulate and analyse our data 
install.packages("ggplot2") # We'll use this package to visualise our data
```

Once installed, load them with the following code:

```{r load libraries}
# Load the libraries
library(dplyr)
library(ggplot2)
```


## Importing the data

With our working directory set, we can now import some data. Today our data will be in `csv` (**C**omma, **S**eparated **V**alues) format, which is a standard format for storing a table in a simple text form.

Here is some code to read in a csv file:
```{r reading in data in csv format}
completeMotorCarTrends <- read.csv("Data/carInfo-Prices_27-05-20.csv", header= TRUE)
```
<!-- [THIS IS A COMMENT] Hugo - worth implementing a naming conventions so that files and folders are informatively named and dated -->

Or we can use the following code to ask you to select a file:
```{r selecting csv data file by user, eval=FALSE}
motorCarTrends <- read.csv(file.choose(), header=TRUE)
```


## Visualising the Data

### Lollipop Chart
Lollipop charts conveys the same information as in bar charts. By reducing the thick bars into thin lines, it reduces the clutter and lays more emphasis on the value. It looks nice and modern. 

Do citizens spend more on their vehicles as they advance in years?   
```{r Lollipop chart}
#Prepare data
groupedCompleteMotorCarTrends<- group_by(completeMotorCarTrends, AgeCategories)
summariseCompleteMotorCarTrends<- summarise(groupedCompleteMotorCarTrends, mean_carModel= mean(CostOfCar))

# Plot
ggplot(summariseCompleteMotorCarTrends, aes(x=AgeCategories, y=mean_carModel)) +
  geom_point(size=3) + geom_segment(aes(x=AgeCategories,
                                        xend=AgeCategories,
                                        y=0,
                                        yend=mean_carModel)) +
  labs(title="Avg. Cost in USDs spent on Car", caption="source: mpg") +
  theme(axis.text.x = element_text(angle=65, vjust= 0.6))
```

### Grouped Bar Chart
Grouped bar charts are used when two or more data sets are displayed side-by-side and grouped together under categories on the same axis. It is essentially a simple bar chart with two or more graphs. 

Are men spending more on vehicles then women?
```{r Grouped bar chart}
#Prepare data
costOfCarMeans<- completeMotorCarTrends %>% group_by(GenderOfBuyer, AgeCategories) %>%
  summarise(mean= mean(CostOfCar))

#Plot
ggplot(costOfCarMeans, aes(x=AgeCategories , y=mean, fill=GenderOfBuyer)) + 
  geom_bar(stat= "identity", position= "dodge") +
  labs(title="Avg. Cost in USDs spent on Car by Gender") 
```

### Ordered Bar Chart
Ordered Bar Chart is a Bar Chart that is ordered by the Y axis variable. Just sorting the dataframe by the variable of interest isn’t enough to order the bar chart. In order for the bar chart to retain the order of the rows, the X axis variable (i.e. the categories) has to be converted into a factor.

What Car Model consumes the least fuel?  
```{r Ordered bar chart}
#Prepare data
milesPerGallonCars<- completeMotorCarTrends %>% group_by(CarModel) %>%
  summarise(mean= mean(MilesPerGallon))

newColumnNames <- c("Car Model","Av. Miles per Gallon")
colnames(milesPerGallonCars) <- newColumnNames
head(milesPerGallonCars)

milesPerGallonCars<- milesPerGallonCars[order(milesPerGallonCars$`Av. Miles per Gallon`), ] #sort
milesPerGallonCars$`Car Model`<- factor(milesPerGallonCars$`Car Model`, levels = milesPerGallonCars$`Car Model`) #retain order in plot

#Plot
ggplot(milesPerGallonCars, aes(x=`Car Model`, y=`Av. Miles per Gallon`)) +
  geom_bar(stat="identity", width=.5, fill="blue") + labs(title="Average number of Miles Car is able to Travel per Gallon") +
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) 
```
### Horizontal Bar Chart
Horizontal bar chart is a slight variation on the vertical bar chart, with the primary difference being the inverted positions of the x-axis and y-axis. To invert the positions of the x and y axis, add the 'coord_flip()' function. The Ordered Bar Chart above has been made into a horizontal bar chart below. 

```{r Ordered horziontal bar chart}
#Create horizontal bar chart
ggplot(milesPerGallonCars, aes(x=`Car Model`, y=`Av. Miles per Gallon`)) +
  geom_bar(stat="identity", width=.5, fill="blue") + labs(title="Average number of Miles Car is able to Travel per Gallon") + coord_flip()  
```

### Boxplot
Box plot is an excellent tool to study the distribution. It can also show the distributions within multiple groups, along with the median, range and outliers if any. The dark line inside the box represents the median. The top of box is 75%ile and bottom of box is 25%ile. The end points of the lines (aka whiskers) is at a distance of 1.5*IQR, where IQR or Inter Quartile Range is the distance between 25th and 75th percentiles. The points outside the whiskers are marked as dots and are normally considered as extreme points.

Is the cost of fuel for car models distributed evenly?  
```{r Boxplot}
#Prepare the data
costPerMileBoxPlot <- completeMotorCarTrends[ , c("CarModel", "CostPerMile")] 
head(costPerMileBoxPlot)

colClasses <- sapply(costPerMileBoxPlot, FUN=class)
print(colClasses)

# Create dataset of vehicle types
Sedan <- c("Cadillac", "Mercedes", "Hornet", "Mazda", "Datsun", "Toyota", "Volvo")
Sports <- c("Maserati", "Porsche", "Ferrari", "Lotus")
Pickup<- "Ford"
Compact<- c("Fiat", "Fiat", "Honda")

costPerMileBoxPlot$CarModel[costPerMileBoxPlot$CarModel %in% Sedan] <- "Sedan"
costPerMileBoxPlot$CarModel[costPerMileBoxPlot$CarModel %in% Sports] <- "Sports"
costPerMileBoxPlot$CarModel[costPerMileBoxPlot$CarModel %in% Pickup] <- "Pickup"
costPerMileBoxPlot$CarModel[costPerMileBoxPlot$CarModel %in% Compact] <- "Compact"

# Plot
Plot<- ggplot(costPerMileBoxPlot, aes(CarModel, CostPerMile))
Plot + geom_boxplot(varwidth = T, fill= "plum") +
  labs(title= "Cost per Mile grouped by Model of Vehicle",
       caption= "Source: VNSO",
       x= "Model of Vehicle",
       y= "Cost per Mile")
```

### Scatter Plot
Scatter plot is a graph in which values of two variables are plotted along two axes, the pattern of the resulting points reveals if any correlation is present (relationship between two variables). It is the most frequently used plot for data analysis. 

Is there 
```{r Scatter plot}
plot(x=completeMotorCarTrends$CostOfCar, y=completeMotorCarTrends$CostPerMile,
     bty="n", # Remove box from around plot
     pch=19, # Set the plotting shape
     las=1, # Set the angle of tick labels to be horizontal
     main="Does spending more on your vehicle reduce your cost on fuel?", # Add a plot title
     ylab="Cost per mile (USDs)", # Y axis label
     xlab="Cost of Vehicle", # X axis label
     col="blue", # Set colour of points
     cex=1.5) # Set the size of the points

# Run a simple linear regression to test whether there is a relationship
linearModelResults <- lm(CostPerMile ~ CostOfCar, data=completeMotorCarTrends)
linearModelResults

# Plotting regression line
abline(lm(CostPerMile ~ CostOfCar))
```

Is there a relationship between horsepower and the time it takes to travel a quarter of a mile?
```{r plotting horsepower versus quarter mile travel time 1}
plot(x=completeMotorCarTrends$HorsePower, y=completeMotorCarTrends$TimeToQuarterMile,
     bty="n", # Remove box from around plot
     pch=19, # Set the plotting shape
     las=1, # Set the angle of tick labels to be horizontal
     main="Does horse power affect speed?", # Add a plot title
     ylab="Time to travel a quarter of a mile (seconds)", # Y axis label
     xlab="Gross horse power", # X axis label
     col="blue", # Set colour of points
     cex=1.5) # Set the size of the points

# Run a simple linear regression to test whether there is a relationship
linearModelResults <- lm(TimeToQuarterMile ~ HorsePower, data=completeMotorCarTrends)

# Replace NA from the column HorsePower with the Mean HorsePower]

completeMotorCarTrends$HorsePower[is.na(completeMotorCarTrends$HorsePower)] = 147

# Generate some example horse power values - for predicting speed
horsePowerValues <- seq(min(completeMotorCarTrends$HorsePower), max(completeMotorCarTrends$HorsePower), by=1)

# Predict the time taken to travel a quarter mile for the example values
confidenceIntervals <- predict(linearModelResults, newdata=data.frame(HorsePower=horsePowerValues),
                               interval="confidence", # Record the confidence intervals around predicted values
                               level=0.95) # Set the level for the confidence intervals

# Add the confidence intervals onto the plot as a polygon
polygon(x=c(horsePowerValues, rev(horsePowerValues)),
        y=c(confidenceIntervals[, 2], rev(confidenceIntervals[,3])), 
        col=rgb(1,0,0, 0.25), # Set the colour to be red and slightly transparent
        border=NA) # Remove the border from around the polygon

# Add a line of best fit to the plot
lines(x=horsePowerValues, 
      y=confidenceIntervals[, 1],
      col="red", # Set the line colour to be red
      lty=2, # Set the type of the line to be dashed
      lwd=2) # Set the width of the line
```
