---
title: "Getting started with Rmarkdown"
author: "Vanuatu National Statistics Office"
date: "25/05/2021"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Introduction

## First part of the introduction

A report with some examples of the historic trades statistics data for **Vanuatu**. To work with the data we will load these required libraries:

```{r preparation, message = FALSE}
# Load the required libraries
library(dplyr) # Manipulating data - install.packages("dplyr")
library(ggplot2) # Plotting data - install.packages("ggplot2")
library(DT) # Creating formatted html tables
```

## Second part of the introduction

Here is an example plot:

### An example of creating a plot

```{r create simple plot}
example_data <- data.frame("Value" = rnorm(500))

example_data %>% 
  ggplot() +
    geom_histogram(aes(x = Value), bins = 10)
```


# Viewing the data

The trade statistics data can be read in with the following code:

```{r load data}
# Read in some data
historic_exports_file <- file.path("..", "..", "vnso-RAP-tradeStats-materials", "data", "secure", "historical_export_99_19.csv")
historic_exports <- read.csv(historic_exports_file)

# View the data
historic_exports[1:25, ] %>% datatable(options = list(pageLength = 10))
```

# Summarising the data

We can generate summaries of the data with this code:

```{r generate summary statistics}
# Calculate summary statistics of Value by year
historic_exports_by_year <-
  historic_exports %>%
  filter(is.na(Value) == FALSE) %>%
  group_by(Year) %>%
  summarise(mean = mean(Value),
            min = min(Value),
            max = max(Value),
            sum = sum(Value),
            median = median(Value), 
            .groups = "drop")
```

Which creates:

```{r view summary statistics}
historic_exports_by_year %>% datatable()
```

# Plotting the data
HGere is a plot of the total exports by year:

```{r plot exports by year}
# Create a plot of median value by year
historic_exports_by_year %>%
  ggplot() +
    aes(x = Year, y = median) +
    geom_col(alpha = 0.5) +
    geom_line(color = "red", linetype = "dashed", size = 1.5) +
    geom_point(color = "blue") +
    labs(title = "Statistic value of exports",
         x ="Year", y = "Value")
```

